Tracking $p$-adic precision

### 2014-05-14 padicprec.tex
First pass fixing typos pointed out by referee

Two sources of inspiration arise when studying $p$-adic algorithms.
The first relates $\Zp$ to its quotients $\Z/p^n\Z$.  The preimage in
$\Zp$ of an element $a \in \Z / p^n \Z$ is a ball, and these balls cover
$\Zp$ for any fixed $n$.  Since the projection $\Zp \to \Z/p^n\Z$ is a
homomorphism, given unknown elements in two such balls we can
locate the balls in which their sum and product lie.  Working on a computer
we must find a way to write elements using only a finite amount of data.
By lumping elements together into these balls of radius $p^{-n}$, we may
model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In this representation,
all $p$-adic elements in have constant \emph{absolute precision} $n$.

@@

The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
\[
u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.
\]
We shall consider the case where the initial
terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.
Let us first examine how the absolute precision of $u_n$ varies with $n$
if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
The computation of $u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
decreases the precision by a factor $p^{\val(u_n)}$.  Hence the
step-by-step computation returns the value of $u_n$ with precision
\begin{equation}
\label{eq:SOMOS}
O(p^{N-v_n})
\quad \text{with}\quad
v_n = \val(u_0) + \cdots + \val(u_{n-4}).
\end{equation}

@@

Let us first recall that a square matrix $M$ with coefficients in $K$ 
admits a LU factorization if it can be written as a product $LU$ where 
$L$ and $U$ is lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems, and is
discussed further in Appendix \ref{ssec:matrices}.
When computing the entries of $L$ and $U$ from a $d \times d$ matrix
over $\Zp$ with entries of precision $O(p^N)$, one has a choice of algorithms:
\begin{itemize}
\item using usual Gaussian elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulae yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$ \cite{caruso:12a}.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. On the other hand, the second algorithm
is less efficient than the first one because evaluating Cramer-type
formulae requires many computations. 
%In \cite{caruso:12a}, the author designs a new algorithm which combines 
%both advantages: efficiency and stability.
%The precision loss inherent in an algorithm is an important factor in 
%analyzing its performance.
%However, since this precision loss depends heavily on the details of the 
%problem and the algorithm, we will not focus on such questions in this 
%paper.

@@

Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
We assume that the induced metric is an ultrametric (\emph{i.e.}
$\lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert)$)
and that $K$ is complete with respect to it.
For example, we may take $K = \Qp$ with the
$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
that $K$ contains a dense subring $R \subset K$
consisting of elements that can be represented with a finite amount of space.
For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
for $K = \Fp\lb t \rb$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.

@@

\begin{deftn} \label{deftn : diff}
Let $E$ and $F$ be two $K$-Banach spaces, let $U$ be an open 
subset of $E$ and let $f : U \rightarrow F$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a continuous linear 
map $f'(v_0) : U \rightarrow W$ such that for any $\varepsilon >0$, 
there exists an open neighborhood $U_\varepsilon \subset U$ containing 
$v_0$ with
\[ 
\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
for all $v, w \in U_\varepsilon$.  The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

@@

\begin{deftn}
\label{def:locanal}
Let $E$ and $F$ be $K$-Banach spaces. Let $U$ be an open subset of $E$
and let $x \in U$.
A function $f \col U \rightarrow F$ is said \emph{locally analytic} at 
$x$ if there exists an open subset $U_x \subset E$ and continuous $n$-linear
maps $L_n \col E^n \rightarrow F$ for $n \ge 1$ such that
\[
f(x+h) = f(x) + \sum_{n \geq 1} L_n(h,\ldots,h)
\]
for all $h$ with $x + h \in U_x$.
\end{deftn}

@@

We now study the effect of truncation on series: given $f$ as above
and a nonnegative integer $n_0$, we set
\[
f_{\geq n_0} = \sum_{n \geq n_0} f_n = f - (f_0 + f_1 + \cdots +
f_{n_0 - 1}).
\]
On the other hand, given a convex function $\varphi : \R \to \R \cup
\{+\infty\}$ and a real number $v$, we define $\varphi_{\geq v} : \R \to 
\R \cup \{\pm \infty\}$ as the highest convex function such that 
$\varphi_{\geq v} \leq \varphi$ and the function $x \mapsto 
\varphi_{\geq v}(x) - v x$ is nondecreasing. Concretely, we have:
$$\varphi_{\geq v}(x) = \inf_{y \geq 0} \, \big(\varphi(x + y) - v y 
\big).$$
When $v$ is fixed, the construction $\varphi \mapsto \varphi_{\geq v}$
is nondecreasing: if $\varphi$ and $\psi$ are two convex functions such
that $\varphi \leq \psi$, we deduce $\varphi_{\geq v} \leq \psi_{\geq v}$.

@@

The execution of the function {\tt f} yields a factorization:
$$f = f_n \circ f_{n-1} \circ \cdots \circ f_1$$
where the $f_i$'s correspond to each individual basic step (like 
addition, multiplication or creation of variables); they are then
``nice'' (in particular locally analytic) functions. For all 
$i$, let $U_i$ denote the codomain of $f_i$. Of course $U_i$ must
contains all possible values of all variables which are defined in the 
program after the execution of $i$-th step. Mathematically, we assume
that it is an 
open subset in some $K$-Banach space $E_i$. We have $U_n = V$ and the 
domain of $f_i$ is $U_{i-1}$ where, by convention, we have set $U_0 = 
U$.
For all $i$, we set $g_i = f_i \circ \cdots \circ f_1$ and $x_i = 
g_i(x)$.

@@

The previous approach works only if the $g'_i(x)$'s are all surjective. 
Unfortunately, this assumption is in general not fulfilled. Indeed, 
remember that the dimension of $E_i$ is roughly the number of used 
variables after the step $i$. It all $g_i'(x)$ were surjective, this 
would mean that the function {\tt f} never initializes a new variable!
In what follows, we propose another solution that does not assume the
surjectivity of $g'_i(x)$.

For $i \in \{1, \ldots, n\}$, define $h_i = f_n \circ \cdots \circ 
f_{i+1}$, so that we have $f = h_i \circ g_i$. On differentials, we 
have $f'(x) = h_i'(x_i) \circ g'_i(x)$. Since $f'(x)$ is surjective (by 
assumption), we deduce that $h'_i(x_i)$ is surjective for all $i$. Let $H'_i$ 
be a lattice in $E_i$ such that:
\begin{enumerate}[(a)] 
\item \label{item:Hi1}
$H'_i$ is contained in $H_i + \ker h'_i(x_i) = h'_i(x_i)^{-1}
\big(f'(x)(H)\big)$;
\item \label{item:Hi2}
$H'_i$ is a first order lattice for $h_i$ at $x_i$.
\end{enumerate}
By definition, we have
$h_i(x_i + H'_i) = x_n + h'_i(x_i)(H'_i) \subset x_n + f'(x)(H)$.
Therefore, modifying the intermediate value $x_i$ by an element
of $H'_i$ after the $i$-th step of the execution of {\tt f} leaves
the final result remains unchanged. In other words, it is
enough to compute $x_i$ modulo $H'_i$.

It is nevertheless not obvious to implement these ideas in practice
because when we enter in the $i$-th step of the execution of {\tt f},
we have not computed $h_i$ yet and hence are \emph{a priori} not able
to determine a lattice $H'_i$ satisfying the axioms \eqref{item:Hi1}
and \eqref{item:Hi2} above.
A possible solution to tackle this problem is to proceed in several
stages as follows: 
\begin{enumerate}[(1)]
\item \label{item:smallprec}
for $i$ from $1$ to $n$, we compute $x_i$, $f'_i(x_{i-1})$ 
at small precision (but enough for the second step) together with an
upper bound of the function $\Lambda(h \mapsto f_i(x_{i-1}+h) - 
f_i(x_{i-1}))$;
\item \label{item:determineHi}
for $i$ from $n$ to $1$, we compute $h'_i(x_i)$ and
determine a lattice $H'_i$ satisfying \eqref{item:Hi1} and 
\eqref{item:Hi2};
\item \label{item:finalcomp}
for $i$ from $1$ to $n$, we recompute $x_i$ modulo $H'_i$
and finally outputs $x_n + O\big(f'(x)(H)\big)$.
\end{enumerate}
Using relaxed algorithms for computing with elements in $K$ (\emph{cf} 
\cites{hoeven:02a, hoeven:07a, berthomieu-hoeven-lecerf:11a}), we can reuse in 
Step~\eqref{item:finalcomp} the computations already performed in 
Step~\eqref{item:smallprec}. The two-pass method we have just 
presented is then probably not much more expansive than the one-pass 
method, although it is more difficult to implement.

@@

The study of the this example actually suggests another solution to 
tackle the issue of non-surjectivity. Indeed, remark that our $f$ above 
had not a surjective differential simply because its codomain was too 
large: if we had replaced $f : (K^n)^n \to M_n(K)$ by $g : (K^n)^n \to 
S_n(K)$ (where $S_n(K)$ denotes the $K$-vector space of symmetric matrix 
over $K$ of size $n$) defined in the same way, our problem would have 
disappeared. Of course the image of a general $f$ is rarely a sub vector 
space of $F$ but it is often a sub-$K$-manifold of $F$ (see Appendix
\ref{sec:manifold}). We can then use 
the results of Appendix \ref{sec:manifold} to study $f$ viewed as a 
function whose codomain is $f(U)$, understood that the differential of 
it has now good probability to be surjective.

@@

\subsubsection*{Quick comparison with floating point arithmetics.}

The two strategies described above share some similarities with usual 
floating point arithmetics over the reals. Indeed, roughly speaking, in 
each setting, we begin by choosing a large precision, we do all our 
computations up to this precision understood that when we are not sure 
about some digit, we choose it ``at random'' or using good heuristics. 
The main difference is that, in the ultrametric setting, we are able 
(under some mild hypothesis) to quantify the precision we need at each 
individual step in order to be sure that the final result is correct up 
to the required precision.

@@

Recall from \S \ref{ssec:stepbystep} that a SOMOS 4 sequence is a 
four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 
+ u_{n+3}^3}{u_n}$. We recall also that SOMOS sequences exhibit the 
Laurent phenomenon: it means that, if the four initial terms $u_0, u_1, 
u_2, u_3$ are some indeterminates, then each $u_n$ is a Laurent 
polynomial with coefficients in $\Z$ in these indeterminates
(see \cite{fomin-zelevinsky:02a}).
From now, we will always consider SOMOS sequences with values in $\Q_p$ 
(for some prime number $p$). We assume for simplicity that $u_0, u_1, u_2, 
u_3$ are all units in $\Z_p$. By the Laurent phenomenon, this implies 
that all $u_n$'s lie in $\Z_p$, and that if $u_0, u_1, u_2, u_3$ 
are known with finite precision $O(p^N)$ then all $u_n$ are 
known with the same absolute precision. Algorithm \ref{algo:SOMOS} presented on 
page \pageref{algo:SOMOS} performs this computation.

>>>>

Two sources of inspiration arise when studying $p$-adic algorithms.
The first relates $\Zp$ to its quotients $\Z/p^n\Z$.  The preimage in
$\Zp$ of an element $a \in \Z / p^n \Z$ is a ball, and these balls cover
$\Zp$ for any fixed $n$.  Since the projection $\Zp \to \Z/p^n\Z$ is a
homomorphism, given unknown elements in two such balls we can
locate the balls in which their sum and product lie.  Working on a computer
we must find a way to write elements using only a finite amount of data.
By lumping elements together into these balls of radius $p^{-n}$, we may
model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In this representation,
all $p$-adic elements have constant \emph{absolute precision} $n$.

@@

The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
\[
u_n = \frac{u_{n-3} u_{n-1} + u_{n-2}^2}{u_{n-4}}.
\]
We shall consider the case where the initial
terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.
Let us first examine how the absolute precision of $u_n$ varies with $n$
if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
The computation of $u_n$ involves a division by $u_{n-4}$ and hence, roughly speaking,
decreases the precision by a factor $p^{\val(u_{n-4})}$.  Hence the
step-by-step computation returns the value of $u_n$ with precision
\begin{equation}
\label{eq:SOMOS}
O(p^{N-v_n})
\quad \text{with}\quad
v_n = \val(u_0) + \cdots + \val(u_{n-4}).
\end{equation}

@@

Let us first recall that a square matrix $M$
admits a LU factorization if it can be written as a product $LU$ where 
$L$ and $U$ are lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems, and is
discussed further in Appendix \ref{ssec:matrices}.
When computing the entries of $L$ and $U$ from a $d \times d$ matrix
over $\Zp$ with entries of precision $O(p^N)$, one has a choice of algorithms:
\begin{itemize}
\item using usual Gaussian elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulae yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$ \cite{caruso:12a}.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. On the other hand, the second algorithm
is less efficient than the first one because evaluating Cramer-type
formulae requires many computations. 
%In \cite{caruso:12a}, the author designs a new algorithm which combines 
%both advantages: efficiency and stability.
%The precision loss inherent in an algorithm is an important factor in 
%analyzing its performance.
%However, since this precision loss depends heavily on the details of the 
%problem and the algorithm, we will not focus on such questions in this 
%paper.

@@

Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
We assume that the induced metric is an ultrametric (\emph{i.e.}
$\lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert)$)
and that $K$ is complete with respect to it.
For example, we may take $K = \Qp$ with the
$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
that $K$ contains a dense subring $R \subset K$
consisting of elements that can be represented with a finite amount of space.
For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
for $K = \Fp\lp t \rp$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.

@@

\begin{deftn} \label{deftn : diff}
Let $E$ and $F$ be two $K$-Banach spaces, let $U$ be an open 
subset of $E$ and let $f : U \rightarrow F$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a continuous linear 
map $f'(v_0) : U \rightarrow W$ such that for any $\varepsilon >0$, 
there exists an open neighborhood $U_\varepsilon \subset U$ containing 
$v_0$ with
\[ 
\Vert f(v)-f(w)-f'(v_0) \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
for all $v, w \in U_\varepsilon$.  The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

@@

\begin{deftn}
\label{def:locanal}
Let $E$ and $F$ be $K$-Banach spaces. Let $U$ be an open subset of $E$
and let $x \in U$.
A function $f \col U \rightarrow F$ is said to be \emph{locally analytic} at 
$x$ if there exists an open subset $U_x \subset E$ and continuous $n$-linear
maps $L_n \col E^n \rightarrow F$ for $n \ge 1$ such that
\[
f(x+h) = f(x) + \sum_{n \geq 1} L_n(h,\ldots,h)
\]
for all $h$ with $x + h \in U_x$.
\end{deftn}

@@

We now study the effect of truncation on series: given $f$ as above
and a nonnegative integer $n_0$, we set
\[
f_{\geq n_0} = \sum_{n \geq n_0} f_n = f - (f_0 + f_1 + \cdots +
f_{n_0 - 1}).
\]
On the other hand, given a convex function $\varphi : \R \to \R \cup
\{+\infty\}$ and a real number $v$, we define $\varphi_{\geq v} : \R \to 
\R \cup \{\pm \infty\}$ by
\[
\varphi_{\geq v}(x) = \inf_{y \geq 0} \, \big(\varphi(x + y) - v y \big).
\]
The function $\varphi_{\geq v}$ is the maximum among all functions $\varphi'$
with $\varphi' \leq \varphi$ and $x \mapsto \varphi'(x) - v x$ nondecreasing.
When $v$ is fixed, the construction $\varphi \mapsto \varphi_{\geq v}$
is nondecreasing: if $\varphi$ and $\psi$ are two convex functions such
that $\varphi \leq \psi$ then $\varphi_{\geq v} \leq \psi_{\geq v}$.

@@

The execution of the function {\tt f} yields a factorization:
$$f = f_n \circ f_{n-1} \circ \cdots \circ f_1$$
where the $f_i$'s correspond to each individual basic step (like 
addition, multiplication or creation of variables); they are then
``nice'' (in particular locally analytic) functions. For all 
$i$, let $U_i$ denote the codomain of $f_i$. Of course $U_i$ must
contains all possible values of all variables which are defined in the 
program after the execution of the $i$-th step. Mathematically, we assume
that it is an 
open subset in some $K$-Banach space $E_i$. We have $U_n = V$ and the 
domain of $f_i$ is $U_{i-1}$ where, by convention, we have set $U_0 = 
U$.
For all $i$, we set $g_i = f_i \circ \cdots \circ f_1$ and $x_i = 
g_i(x)$.

@@

The previous approach works only if the $g'_i(x)$'s are all surjective. 
Unfortunately, this assumption is in general not fulfilled. Indeed, 
remember that the dimension of $E_i$ is roughly the number of used 
variables after the step $i$. If all $g_i'(x)$ were surjective, this 
would mean that the function {\tt f} never initializes a new variable!
In what follows, we propose another solution that does not assume the
surjectivity of $g'_i(x)$.

For $i \in \{1, \ldots, n\}$, define $h_i = f_n \circ \cdots \circ 
f_{i+1}$, so that we have $f = h_i \circ g_i$. On differentials, we 
have $f'(x) = h_i'(x_i) \circ g'_i(x)$. Since $f'(x)$ is surjective (by 
assumption), we deduce that $h'_i(x_i)$ is surjective for all $i$. Let $H'_i$ 
be a lattice in $E_i$ such that:
\begin{enumerate}[(a)] 
\item \label{item:Hi1}
$H'_i$ is contained in $H_i + \ker h'_i(x_i) = h'_i(x_i)^{-1}
\big(f'(x)(H)\big)$;
\item \label{item:Hi2}
$H'_i$ is a first order lattice for $h_i$ at $x_i$.
\end{enumerate}
By definition, we have
$h_i(x_i + H'_i) = x_n + h'_i(x_i)(H'_i) \subset x_n + f'(x)(H)$.
Therefore, modifying the intermediate value $x_i$ by an element
of $H'_i$ after the $i$-th step of the execution of {\tt f} leaves
the final result unchanged. In other words, it is
enough to compute $x_i$ modulo $H'_i$.

It is nevertheless not obvious to implement these ideas in practice
because when we enter in the $i$-th step of the execution of {\tt f},
we have not computed $h_i$ yet and hence are \emph{a priori} not able
to determine a lattice $H'_i$ satisfying the axioms \eqref{item:Hi1}
and \eqref{item:Hi2} above.
A possible solution to tackle this problem is to proceed in several
stages as follows: 
\begin{enumerate}[(1)]
\item \label{item:smallprec}
for $i$ from $1$ to $n$, we compute $x_i$, $f'_i(x_{i-1})$ 
at small precision (but enough for the second step) together with an
upper bound of the function $\Lambda(h \mapsto f_i(x_{i-1}+h) - 
f_i(x_{i-1}))$;
\item \label{item:determineHi}
for $i$ from $n$ to $1$, we compute $h'_i(x_i)$ and
determine a lattice $H'_i$ satisfying \eqref{item:Hi1} and 
\eqref{item:Hi2};
\item \label{item:finalcomp}
for $i$ from $1$ to $n$, we recompute $x_i$ modulo $H'_i$
and finally outputs $x_n + O\big(f'(x)(H)\big)$.
\end{enumerate}
Using relaxed algorithms for computing with elements in $K$ (\emph{cf} 
\cites{hoeven:02a, hoeven:07a, berthomieu-hoeven-lecerf:11a}), we can reuse in 
Step~\eqref{item:finalcomp} the computations already performed in 
Step~\eqref{item:smallprec}. The two-pass method we have just 
presented is then probably not much more expensive than the one-pass 
method, although it is more difficult to implement.

@@

The study of this example actually suggests another solution to 
tackle the issue of non-surjectivity. Indeed, we remark that our $f$ above 
did not have a surjective differential simply because its codomain was too 
large: replacing the codomain with the $K$-vector space $S_n(K)$
of $n \times n$ symmetric matrices over $K$ makes the differential of $f$ surjective.
While the image of a general $f$ is rarely a sub vector 
space of $F$, it is often a sub-$K$-manifold of $F$. We can then use 
the results of Appendix \ref{sec:manifold} to study $f : U \to f(U)$, which
likely has surjective differential.

@@

\subsubsection*{Quick comparison with floating point arithmetic.}

The two strategies described above share some similarities with standard 
floating point arithmetic over the reals. In each setting, we begin
by choosing a large precision for all computations, and when we
encounter an unconstrained digit we choose it ``at random'' or using good heuristics. 
However, in the ultrametric setting, mild hypotheses allow us
to quantify the precision needed at each 
individual step in order to ensure a specified final precision.

@@

Recall from \S \ref{ssec:stepbystep} that a SOMOS 4 sequence is a 
four-term inductive sequence defined by $u_n = \frac{u_{n-3} u_{n-1} 
+ u_{n-2}^2}{u_n}$ exhibiting the Laurent phenomenon.
We will focus on SOMOS sequences with values in $\Q_p$, and 
assume for simplicity that $u_0, u_1, u_2, u_3 \in \Z_p^\times$.
By the Laurent phenomenon, $u_n \in \Z_p$ for all $n$, and if $u_0, u_1, u_2, u_3$ 
are known with finite precision $O(p^N)$ then all $u_n$ are 
known with the same absolute precision. Algorithm \ref{algo:SOMOS} presented on 
page \pageref{algo:SOMOS} performs this computation.
